Boosted Mirror Descent, Frank-Wolfe, and Herding

### 2013-09-09 icml/paper.tex
added entropy material and discussion; checked proofs of infinite dimensional issues (none found); added a bunch of comments

@@

In this document we consider a family of methods for minimizing $L(u)$, 
which are based on Bach's generalization of the Frank-Wolfe algorithm and can 
be interpreted as boosted mirror descent.

First, let us generalize the setting 
to consider a \emph{two-argument} loss function
\[ L(u,\theta) = h(u) + \langle \theta, u \rangle - R(\theta). \]
In the case $h(u) \equiv 0$, we can get the previous setting by 
considering $\max_{\theta \in \Theta} L(u,\theta)$. We will assume 
throughout this document that $h$ and $R$ are both convex functions. \NA{JH: Should we comment on the symmetry introduced by this formulation?}

Boosted mirror descent can be thought of as an algorithm for finding 
``saddle points'' of $L$. It is the following algorithm:
\begin{enumerate}
\item $u_1 = 0$
\item $\theta_{t} \in \arg\max_{\theta \in \Theta} \langle \theta, u_t \rangle - R(\theta)$
\item $u_{t+1} \in \arg\min_{u} h(u) + \langle \frac{1}{t} \sum_{s \leq t} \theta_s, u \rangle$
\end{enumerate}
Let $\hat{u} = \frac{1}{T} \sum_{t = 1}^{T} u_t$. As long as $h$ is strongly convex, we obtain the 
bound:
\[ \sup_{\theta \in \Theta} L(\hat{u}, \theta) \leq \sup_{\theta \in \Theta} L(u^*, \theta) + O(\log(T)/T). \]
However, it is often the case that $h$ is not strongly convex, whereas $R$ is strongly convex. In this case 
we may wish to use the following slightly different algorithm:
\begin{enumerate}
\item $\theta_1 = 0$
\item $u_t \in \arg\min_{u} h(u) + \langle \theta_t, u \rangle$
\item $\theta_{t+1} \in \arg\max_{\theta \in \Theta} \langle \theta, \frac{1}{t} \sum_{s \leq t} u_s \rangle - R(\theta)$
\end{enumerate}
Once again, if we let $\hat{u} = \frac{1}{T} \sum_{t=1}^T u_t$, then we obtain the following bound, which now 
only requires the convexity of $R$:
\[ \sup_{\theta \in \Theta} L(\hat{u}, \theta) \leq \sup_{\theta \in \Theta} L(u^*, \theta) + O(\log(T)/T). \]
\NA{JH: It would be more elegant to argue either here or below that the bounds on the second algorithm follow directly from duality}

@@

be chosen via $x_{t+1} \in \arg\min_{x} \frac{1}{t} \sum_{s \leq t} l_s(x)$. 

@@

@@

for all $x^*$, which is true because $x_2 \in \arg\min_x l_1(x)$ by construction. For 

@@

\label{lem:convexity}

@@

Let $F(y) := \arg\min_x f(x) + y^Tx$. Then $\|F(y_0)-F(y_1)\| \leq \frac{2}{\alpha} \|y_0-y_1\|$.

@@

@@

Consider the updates $\theta_t \in \arg\max_{\theta} \langle \theta, u_t \rangle - R(\theta)$ 
and $u_{t+1} \in \arg\min_u h(u) + \langle \frac{1}{t} \sum_{s \leq t} \theta_s, u \rangle$. 

@@

@@

@@

@@

Consider the updates $u_t \in \arg\min_{u} h(u) + \langle \theta_t, u \rangle$ 
and $\theta_{t+1} \in \arg\max_{\theta} \langle \theta, \frac{1}{t} \sum_{s \leq t} u_s \rangle - R(\theta)$. 

@@

\section{Application to Herding}
\label{sec:herding}
Suppose we are given a family of features $\phi$ together 
with a known value $\bar{\phi}$ for $\bE_{\mu}[\phi]$. We 
would like to construct a distribution that approximately 
matches this distribution, i.e.~one for which 
$\bE_{\hat{\mu}}[\phi] \approx \bar{\phi}$. A natural way 
to do this is by using boosted mirror descent to minimize the 
maximum mean discrepancy relative to $\phi$; 
in other words, to minimize
\[ L(\mu) = \max_{i \in \{1,\dots,d\}} |\bE_{\mu}[\phi_i]-\bar{\phi}_i|. \]
If we wish to write this more similarly to the original problem 
formulation, we can write
\[ L(\mu) = \sup_{\|\theta\|_1 \leq 1} \theta^T(\bE_{\mu}[\phi]-\bar{\phi}). \]
One issue is that $\|\theta\|_1$ is not strongly convex, so we relax $\|\theta\|_1$ 
to $\|\theta\|_2$ and write our loss function in the Lagrangian form:
\[ L(\mu, \theta) = \theta^T\bE_{\mu}[\phi] - \theta^T\bar{\phi} - \frac{1}{2}\|\theta\|_2^2. \]
Intriguingly, Method 1 and Method 2 are identical to each other for this choice of $L$ 
(up to a change of index). In this case, $\mu_t = \delta_{x_t}$ for some $x_t$, and 
the updates are the same as the standard herding updates. Let $R = \sup_{x} \|\phi(x)\|_2$; 
then using the Method 2 convergence bound gives us
\begin{align*}
\sup_{\theta} L\left(\frac{1}{T} \sum_{t=1}^T \delta_{x_t}, \theta\right) &\leq \inf_{u^*} \sup_{\theta} L(u^*, \theta) + \frac{2R^2\log(T+1)}{T} \\
 &= \sup_{\theta} \inf_{u^*} L(u^*, \theta) + \frac{2R^2\log(T+1)}{T} \\
 &= \frac{2R^2\log(T+1)}{T}.
\end{align*}
This is a slightly weaker version of the typical herding bound on MMD.

@@

>>>>

\NA{JHH: I'm thinking that from the beginning we should let $u$ be defined over a different space than $\theta$ and assume a linear map $u \mapsto \eta(u) \in \Theta$. Then write $\langle \theta, \eta(u) \rangle$ so the setup will align with the herding case, where $\eta(u) = \bE_{u}[\phi(x)]$. }

@@

In this paper we consider a family of methods for minimizing $L(u)$, 
which are based on Bach's generalization of the Frank-Wolfe algorithm and can 
be interpreted as boosted mirror descent.

First, let us generalize the setting 
to consider a \emph{two-argument} loss function
\[ L(u,\theta) = h(u) + \langle \theta, u \rangle - R(\theta). \]
In the case $h(u) \equiv 0$, we can get the previous setting by 
considering $\max_{\theta \in \Theta} L(u,\theta)$. We will assume 
throughout this document that $h$ and $R$ are both convex functions. \NA{JHH: Should we comment on the symmetry introduced by this formulation?}

Boosted mirror descent can be thought of as an algorithm for finding 
``saddle points'' of $L$. It is the following algorithm:
\begin{enumerate}
\item $u_1 = \argmin_{u} h(u)$
\item $\theta_{t} \in \argmax_{\theta \in \Theta} \langle \theta, u_t \rangle - R(\theta)$
\item $u_{t+1} \in \argmin_{u} h(u) + \langle \frac{1}{t} \sum_{s \leq t} \theta_s, u \rangle$
\end{enumerate}
Let $\hat{u} = \frac{1}{T} \sum_{t = 1}^{T} u_t$. As long as $h$ is strongly convex, we obtain the 
bound:
\[ \sup_{\theta \in \Theta} L(\hat{u}, \theta) \leq \sup_{\theta \in \Theta} L(u^*, \theta) + O(\log(T)/T). \]
However, it is often the case that $h$ is not strongly convex, whereas $R$ is strongly convex. In this case 
we may wish to use the following slightly different algorithm:
\begin{enumerate}
\item $\theta_1 = 0$
\item $u_t \in \argmin_{u} h(u) + \langle \theta_t, u \rangle$
\item $\theta_{t+1} \in \argmax_{\theta \in \Theta} \langle \theta, \frac{1}{t} \sum_{s \leq t} u_s \rangle - R(\theta)$
\end{enumerate}
Once again, if we let $\hat{u} = \frac{1}{T} \sum_{t=1}^T u_t$, then we obtain the following bound, which now 
only requires the convexity of $R$:
\[ \sup_{\theta \in \Theta} L(\hat{u}, \theta) \leq \sup_{\theta \in \Theta} L(u^*, \theta) + O(\log(T)/T). \]
\NA{JHH: It would be more elegant to argue either here or below that the bounds on the second algorithm follow directly from duality}

@@

be chosen via $x_{t+1} \in \argmin_{x} \frac{1}{t} \sum_{s \leq t} l_s(x)$. 

@@

\NA{JHH: Doesn't rely on the spaces $x$'s having any structure nor on any properties of the functions.}

@@

for all $x^*$, which is true because $x_2 \in \argmin_x l_1(x)$ by construction. For 

@@

\begin{lemma} \label{lem:convexity}

@@

Let $F(y) := \argmin_x f(x) + y^Tx$. Then $\|F(y_0)-F(y_1)\| \leq \frac{2}{\alpha} \|y_0-y_1\|$.

@@

\NA{JHH: Follows from $f$ being strongly convex, the definition of inner product, and Cauchy-Schwartz. We'll just want to be careful about calculating $\alpha$ for our regularizer later on.}

@@

Consider the updates $\theta_t \in \argmax_{\theta} \langle \theta, u_t \rangle - R(\theta)$ 
and $u_{t+1} \in \argmin_u h(u) + \langle \frac{1}{t} \sum_{s \leq t} \theta_s, u \rangle$. 

@@

\NA{JHH: Uses, Lemma~\ref{lem:ftl}, convexity of $h$, taking supremums, and facts that follow from the construction of the algorithm.}

@@

 &\leq \frac{1}{T} \sum_{t=1}^T L(u_t, \theta) & \\

@@

\NA{JHH: Follows from Lemma~\ref{lem:convexity}, triangle inequality, taking supremums, and Cauchy-Schwartz.}

@@

Consider the updates $u_t \in \argmin_{u} h(u) + \langle \theta_t, u \rangle$ 
and $\theta_{t+1} \in \argmax_{\theta} \langle \theta, \frac{1}{t} \sum_{s \leq t} u_s \rangle - R(\theta)$. 

@@

\section{Application to Herding}
\label{sec:herding}

\NA{JHH: Need to specify the spaces we are working with here.}
Suppose we are given a family of features $\phi$ together 
with a known value $\bar{\phi}$ for $\bE_{\mu}[\phi]$. We 
would like to construct a distribution that approximately 
matches this distribution, i.e.~one for which 
$\bE_{\hat{\mu}}[\phi] \approx \bar{\phi}$. A natural way 
to do this is by using boosted mirror descent to minimize the 
maximum mean discrepancy relative to $\phi$; 
in other words, to minimize
\[ L(\mu) = \max_{i \in \{1,\dots,d\}} |\bE_{\mu}[\phi_i]-\bar{\phi}_i|. \]
If we wish to write this more similarly to the original problem 
formulation, we can write
\[ L(\mu) = \sup_{\|\theta\|_1 \leq 1} \theta^T(\bE_{\mu}[\phi]-\bar{\phi}). \]
One issue is that $\|\theta\|_1$ is not strongly convex, so we relax $\|\theta\|_1$ 
to $\|\theta\|_2$ and write our loss function in the Lagrangian form:
\[ L(\mu, \theta) = \theta^T\bE_{\mu}[\phi] - \theta^T\bar{\phi} - \frac{1}{2}\|\theta\|_2^2. \]
Intriguingly, Method 1 and Method 2 are identical to each other for this choice of $L$ 
(up to a change of index). In this case, $\mu_t = \delta_{x_t}$ for some $x_t$, and 
the updates are the same as the standard herding updates. Let $R = \sup_{x} \|\phi(x)\|_2$; 
then using the Method 2 convergence bound gives us
\begin{align*}
\sup_{\theta} L\left(\frac{1}{T} \sum_{t=1}^T \delta_{x_t}, \theta\right) &\leq \inf_{u^*} \sup_{\theta} L(u^*, \theta) + \frac{2R^2\log(T+1)}{T} \\
 &= \sup_{\theta} \inf_{u^*} L(u^*, \theta) + \frac{2R^2\log(T+1)}{T} \\
 &= \frac{2R^2\log(T+1)}{T}.
\end{align*}
This is a slightly weaker version of the typical herding bound on MMD.

@@

Let us now consider the entropic regularizer $h(u) = \tau\bE_{u}[\log u(x)]$ with 
\[ R(\theta) = \bE_{u}[\theta^T\bar{\phi}] + \left\{ \begin{array}{llc} \infty & : & \|\theta\|_1 > 1 \\ 0 & : & \|\theta\|_1 \leq 1 \end{array} \right.\]
If we use Method 1, we then end up with 
\[ u_{t}(x) \propto \exp\left\{-\frac{1}{\alpha}\overline{\theta}_{t}^T\phi(x)\right\}, \]
where $\overline{\theta}_{t} = \frac{1}{t} \sum_{s \le t} \theta_{s}$. Hence, $h(u) + \bE_{\mu}[\theta^T\phi(x)] - R(\theta)$ is no longer maximized at the boundary (when $\tau > 0$). We can instead consider the following sequence of distributions:
\begin{enumerate}
\item $\theta_t \in \argmax_{\|\theta\|_1 \leq 1} \theta^T (\bE_{u_t}[\phi(x)]-\bar{\phi})$
\item $u_{t+1}(x) = Z(\overline{\theta}_{t}, \tau)^{-1} \exp\left\{-\frac{1}{\tau}\overline{\theta}_{t}^T\phi(x)\right\}$
\end{enumerate}
Here $Z(\overline{\theta}_{t}, \tau) = \int\exp\left\{-\frac{1}{\tau}\overline{\theta}_{t}^T\phi(x)\right\} \dee x$.

The entropic regularizer is not strongly convex but a more general analysis involving 
Bregman divergence can nevertheless yield a convergence bound in this case. We thus 
get something that looks like herding but actually gives us an exponential family 
type distribution. In particular, we have a Gibbs distribution with temperature $\tau$, and 
in the limit as the temperature goes to zero, we recover herding, since 
\[ \lim_{\tau \to 0} u_{t+1}(x) = \delta_{x^{*}_{t+1}}, \]
where $x^{*}_{t+1} = \argmin_{x} \overline{\theta}_{t}^T\phi(x)$. 
This fact is reminiscent of Welling's original derivation of herding, in which 
he took the zero temperature limit of the gradient descent algorithm for the optimization problem 
\[ \argmin_{u} h(u) \quad \text{s.t.}~\bE_{u}[\phi(x)] - \bar\phi = 0. \]
This optimization is quite similar to the entropic optimization considered here, which 
we can write as 
\begin{align*}
\argmin_{u}& \argmax_{|\theta\|_1 \leq 1} h(u) + \theta^{T}(\bE_{u}[\phi(x)] - \bar \phi)  \\
&= \argmin_{u} h(u) + \|\bE_{u}[\phi(x)] - \bar\phi\|_{2}. 
\end{align*}
In the entropic formulation, the expectation constraint $\bE_{u}[\phi(x)] - \bar\phi = 0$ becomes
soft, though the two optimizations are equivalent in the zero-temperature limit.

We can view the entropic regularization
with non-zero temperature as an alternative approximation to herding. If at each step we
retain a sample $x_{t} \sim u_{t}$, then the sequence $(x_{t})_{t}$ converges to the 
herding sample path as $\tau \to 0$. So $(x_{t})_{t}$ with non-zero temperature approximate
herding samples. \NA{JHH: I'm not sure how useful this all is, other than to make contact with 
the original paper, as I think the Gibbs distribution was exactly what Max was trying to 
avoid by taking the zero-temperature limit.}
